---
title: "Pokemon `dplyr` practice"
code-annotations: hover
echo: true
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| include: false
library(readr) 
library(dplyr)
library(tidyr) 

poke <- read_csv("https://bit.ly/poke-data") 
poke_types <- separate(poke, type, c("type1", "type2"), sep = ",") 
```

## Filter

Using `poke_types`, get Pokemon matching the following criteria:

1. Pokedex number between 800 and 900
2. Grass type
3. Attack stats of more than 120


```{r}
#| echo: true

poke_types |>
  filter(pokedex_no >= 800, pokedex_no <= 900) |>
  filter(type1 == "Grass") |>
  filter(attack > 120)
```


## Select 

Using `poke_types`, get only columns matching any of the following criteria:


1. columns starting with `t`
2. the 9th column
3. the pokemon name

```{r}
#| echo: true

poke_types |>
  select(starts_with("t"), 9, name)
```


## Mutate 

Create a column named `n_elevator` that calculates how many of these pokemon 
could fit in an elevator with a weight limit of 6000 lbs. 
You may assume that pokemon can be infinitely compressed, so that the volume of the pokemon is not a concern.

Round your answer down to the nearest integer with the `floor()` function - no one needs to get hurt!

Remember, you will need to convert kg to lbs, so you may want to create an additional column named `weight_lbs`.

Output only `gen`, `name`, and the columns you created.

```{r}
#| echo: true

poke_types |>
  mutate(weight_lbs = weight_kg * 2.204623, 
         n_elevator = floor(6000/weight_lbs)) |>
  select(gen, name, weight_lbs, n_elevator)
```

## Group_by + Summarize! 

Work with `poke_types` and only worry about the primary type (`type1`)


### Bug Pokemon by Generation
Calculate the total number of bug pokemon in each generation.

```{r}
#| echo: true

poke_types |>
  filter(type1 == "Bug") |>
  group_by(gen) |>
  summarize(Bug = n())

# OR

poke_types |>
  group_by(gen) |>
  summarize(Bug = sum(type1=="Bug"))
```

### Pokemon Weight By Type

Find the heaviest pokemon of each type, making sure to remove any duplicate rows. 

```{r}
poke_types |>
  select(gen, name, type1, weight_kg) |> # Keep only basic info
  unique() |> # Remove duplicate rows
  group_by(type1) |>
  filter(weight_kg == max(weight_kg))
```

### Pokemon HP by Generation

Find the pokemon with the highest HP for each generation. Include all tied rows.

```{r}
poke_types |>
  select(gen, name, hp) |>
  unique() |>
  group_by(gen) |>
  filter(hp == max(hp)) |>
  arrange(gen)
```

