---
title: "2.1.a - Class Activity"
format: html
number-sections: true
execute:
  echo: true   # show code by default
  warning: false
  message: false
---

This assignment illustrates the important of data visuals when trying to understand the distribution of data. We will use data available in the `datasauRus`. Please be sure you have installed this package (perhaps using the `install.packages()` function) before you try to proceed with the activity. 

```{r}
library(tidyverse)
library(datasauRus)
```

# Problem 

**Discuss with a partner:** What is the difference in the two datasets printed below? Which of the two datasets would be considered "tidy"?

```{r}
datasaurus_dozen_wide
datasaurus_dozen
```

*The "wide" version has each of the different pairs of coordinates in separate columns. This is not tidy, since there should be only one column devoted to "x" coordinates and one column devoted to "y" coordinates. The non wide version of the data is tidy because it only has one column per coordinate and includes an additional column specifying the data type.*

# Problem 

Use a tidyverse chain (code separated by pipes) to determine the mean and standard deviation for the "x" and "y" variable for each unique dataset in the `datasaurus_dozen` tibble (i.e. tidyverse data.frame). This means that your dataset should have four variables: 

- `mean_x`
- `mean_y`
- `sd_x`
- `sd_y`

```{r}
datasaurus_summary <- datasaurus_dozen |> 
  group_by(dataset) |> 
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y)
  )

datasaurus_summary
```

# Problem

**Discuss with a partner:** What do you observe about the grouped means and standard deviations for both variables? 

*They all have nearly identical means and standard deviations for each group of points*. 

# Problem

Create a scatterplot (pairs of x-y points) for all 12 groups in the `datasaurus_dozen` dataset. 

```{r}
ggplot(datasaurus_dozen, aes(x = x, y = y)) + 
  geom_point() + 
  facet_wrap(~ dataset) + 
  theme_bw()
```

Note that each of the above plots follow the principle of "small multiples" which means that each plot is identical in format and differs only in the data that is represented. This is a really powerful way to compare information across groups. 

# Problem 

**Discuss with a partner:** Describe what you see in the above plot. What do these visuals say about the representative-ness of the mean and standard deviation in describing the distribution of the data? 

*The plots are all create unique visuals, including a dinosaur! This suggests that the mean and standard deviation were poor representations of the data and that the only way to truly understand these data were to visualize them.*





