---
title: "2.3.a - Class Activity"
format: html
number-sections: true
execute:
  echo: true   # show code by default
  warning: false
  message: false
---

This assignment is intended to give you experience using `ggplot2`. For this assignment, we will be using the `mpg_hawkes` dataset available in the `stat1080r` package. Recall from the homework assignments that this package is installed using the command `install.packages("stat1080r", repos = "https://beanb2.r-universe.dev")`. Remember that you only need to install this package once per semester. However, you still need to use `library(stat1080r)` each time you start R and want to use it. Do not try to run the following block of code until after the `stat1080r` package has been installed. 

```{r}
library(tidyverse)
library(stat1080r)
```

This assignment will focus on the `mpg_hawkes` dataset. For more information about the variables in `mpg_hawkes`, please use the `?mpg_hawkes` command in the console. Note that the dataset will only be available to use after running `library(stat1080r)`.  

The following code takes `mpg_hawkes` and creates two new variables. The first variable (`decade`) is a factor based on the decade the vehicle was made. The second variable (`v_class_2`) is a factor based on the vehicle type, after using "regular expressions" to group some smaller categories into larger groups. **Run this block once at the start. You don’t need to re-run it unless you restart R.**

```{r}
mpg_hawkes <- mpg_hawkes |>
  mutate(
    decade = case_when(
      year >= 1980 & year < 1990 ~ "1980s",
      year >= 1990 & year < 2000 ~ "1990s",
      year >= 2000 & year < 2010 ~ "2000s",
      year >= 2010 & year < 2020 ~ "2010s"
    ),
    v_class_2 = case_when(
      regexpr("Cars", VClass) > 0 ~ "Cars",
      regexpr("Two Seaters", VClass) > 0 ~ "Cars",
      regexpr("Utility", VClass) > 0 ~ "SUV/Van",
      regexpr("[Vv]an", VClass) > 0 ~ "SUV/Van",
      regexpr("Truck", VClass) > 0 ~ "Trucks",
      TRUE ~ "Other"
    )
  )
```

# Problem

The following block of code uses `dplyr` (i.e., one of the packages in the tidyverse) function to count the number of vehicles of each class. 

```{r}
v_count <- mpg_hawkes |>
  group_by(v_class_2) |> # group by vehicle type
  summarize(n = n()) # assign the counts of each type to the variable "n"

v_count
```

This code shows that there are nearly 20k "Cars" and about 8k "SUV/Van" in the dataset. 

For this problem, copy the code from the above code chunk and adapt it to count the number of vehicles per decade. **Question for Canvas: How many vehicles are included in the dataset from the 1980s?** 

```{r}
# Copy the code from above and adapt as needed.
decade_count <- mpg_hawkes |>
  group_by(decade) |>
  summarize(n = n())

decade_count
```



# Problem

Create a dataset called `mpg_select` that 

- only considers the variables `city`, `highway`, `year`, and `v_class_2`
- only considers vehicles listed as "Trucks" or "SUV/Van" (HINT: remember the `%in%` function)
- only considers vehicles made since the year 2000 (including the year 2000). 

**Question for Canvas: How many vehicles remain in `mpg_select`?**

```{r}
mpg_select <- mpg_hawkes |>
  select(city, highway, year, v_class_2) |>
  filter(year >= 2000 & v_class_2 %in% c("Trucks", "SUV/Van"))
```

# Problem 

Create a histogram of city miles per gallon (MPG) using all vehicles in `mpg_select`. Do not worry about any visual optimization at this stage. 

```{r}
mpg_select |>
  ggplot(aes(x = city)) +
  geom_histogram()
```

# Problem 

Now create an improved version of this histogram by doing the following. 

- use the `binwidth` argument to make each bin of the histogram represent 2 MPG increments. 
- use the `fill` argument OUTSIDE of `aes()` to change the fill of the histogram to be the color "gray50". 
- use the `color` argument OUTSIDE of `aes()` to change the color of the histogram square boundaries to be "black". 
- use `theme_bw()` with no additional arguments to change the plot background. 

Assign your plot to the variable `p1` and then call `p1` by printing it to the screen. 

```{r}
p1 <- mpg_select |>
  ggplot(aes(x = city)) +
  geom_histogram(
    binwidth = 2,
    color = "black", fill = "gray50"
  ) +
  theme_bw()

p1 # Print to the screen
```

**Discuss with a partner:** How would you describe the distribution of city MPG (left-skewed, right-skewed, symmetric) and why? **Note that you will input the skewness type in the Canvas quiz.**

*I would describe the distribution of city MPG as right skewed because there are a few vehicles that have much higher MPG relative to the typical MPG, which tends to be around 15 MPG.*

# Problem 

For this next plot, we are going to add another layer to the `p1` plot. The layer you are going to add is `facet_wrap(~v_class_2, ncol = 1, scale = "free_y")`. Print this updated plot to the screen. The result should be two histograms stacked on top of each other, one showing the City MPG for SUV/Van and the other for Trucks. 

```{r}
p1 + facet_wrap(~v_class_2, ncol = 1, scale = "free_y")
```
# Problem 

The following code shows you how to create colored density plots to compare the city MPG for SUV/Vans and Trucks.

```{r}
mpg_select |>
  ggplot(aes(x = city, color = v_class_2)) +
  geom_density(lwd = 1, bw = 1) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_bw()
```
**Discuss with a partner:** Using the histograms or the density plot, describe the difference in the city MPG for SUV/Van vs Trucks. Also, please note which visual you preferred (histogram vs density plot) for making that comparison. 

*Trucks tend to get worse city MPG compared to SUV/Vans, though both Trucks and SUVs seem to have similar proportions of vehicles with city MPG ratings of 13 MPG or less. SUV/Vans seem more likely to have MPG ratings above 20 MPG compared to trucks.*
