---
title: "7.1.a - Class Activity"
format: html
number-sections: true
execute:
  echo: true   # show code by default
  warning: false
  message: false
---

This assignment is intended to give you experience using `ggplot2` as well as developing an understanding of quantiles in numeric data.
For this assignment, we will be using the `mpg_hawkes` dataset available in the `stat1080r` package. 
You only need to install this package once per semester, using `install.packages("stat1080r", repos = "https://beanb2.r-universe.dev")`. 
However, you still need to use `library(stat1080r)` each time you start R. 
The block of code below will check that `stat1080r` is installed, and if not, it will install it before loading the library. 
It will also check that essential packages within the tidyverse are installed, and if not, will install those packages.

```{r}
#| echo: false
#| include: false
#| label: setup
if(!"stat1080r" %in% installed.packages()) {
  install.packages("stat1080r", repos = "https://beanb2.r-universe.dev")
}
if(!"dplyr" %in% installed.packages()) {
  install.packages("dplyr")
}
if(!"tidyr" %in% installed.packages()) {
  install.packages("tidyr")
}
if(!"ggplot2" %in% installed.packages()) {
  install.packages("ggplot2")
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stat1080r)
```

This assignment will focus on the `mpg_hawkes` dataset. For more information about the variables in `mpg_hawkes`, please use the `?mpg_hawkes` command in the console. Note that the dataset will only be available to use after running `library(stat1080r)` in your R session.  

The following code takes `mpg_hawkes` and creates two new variables. 
The first variable, `decade`, is a **factor** variable that provides the decade the vehicle was made. 
The second variable, `v_class_2`, is a **factor** variable that provides the vehicle type. 
I have used *regular expressions* to group some smaller categories into larger groups. 
**Note that you must run this provided code in order to answer the rest of the questions in this assignment.**

```{r}
#| echo: false
#| include: false
#| label: setup2
mpg_hawkes <- mpg_hawkes |> 
  mutate(decade = case_when(year >= 1980 & year < 1990 ~ "1980s",
                            year >= 1990 & year < 2000 ~ "1990s",
                            year >= 2000 & year < 2010 ~ "2000s",
                            year >= 2010 & year < 2020 ~ "2010s"),
         v_class_2 = case_when(regexpr("Cars", VClass) > 0 ~ "Cars",
                               regexpr("Two Seaters", VClass) > 0 ~ "Cars",
                               regexpr("Utility", VClass) > 0 ~ "SUV/Van",
                               regexpr("[Vv]an", VClass) > 0 ~ "SUV/Van",
                               regexpr("Truck", VClass) > 0 ~ "Trucks",
                               TRUE ~ "Other")
  )
```

# Problem: Types of Variables

The following code considers a subset of the variables in `mpg_hawkes`

```{r}
mpg_select <- mpg_hawkes |>
  select(city, fuelType, trans, year)

head(mpg_select)
```

Of these four variables, which are "quantitative" and which are "categorical" (i.e., "qualitative"), **please provide your answer to this question in the associated Canvas quiz.**

*Replace this text with your answer. Optional if discussing in a group and required if doing this on your own.*

# Problem: Calculating Quantiles

The following code uses the `quantile` function to determine the 90th percentile of city miles per gallon (mpg). This is the mpg quantity for which 90 percent of the 38,000+ vehicles in our dataset have city mpg ratings less than or equal to this value. 

```{r}
q90_city <- mpg_hawkes |>
  pull(city) |> # extract the city column as a vector
  quantile(0.9) # extract the 90th percentile of the city column vector

q90_city
```

Using the above code as reference, find the 90th percentile **highway** mpg rating for the vehicles in our dataset **and then note this amount in your Canvas quiz**.  

```{r}
## q90_highway <- #...#
  
## q90_highway
```

# Problem: Variable Interpretation

**Discuss briefly (2 min or less) with a partner** Based on what you know about vehicles, why do you think that the highway mpg are larger than city mpg? 

*Replace this text with your answer. Optional if discussing in a group and required if doing this on your own.*

# Problem: City and Highway Mileage

Create a plot that compares city (x-axis) vs highway (y-axis) mileage for "Cars" (as defined in the `v_class_2` variable) from the 1980s.

HINT: This will require a filter statement with two logical conditions. 

This will require you to first subset the rows of the data and then use `ggplot2` to visualize. 
The provided code will create the "line of equality" (i.e., the line where x = y) which will help us to visualize how different city and highway mileage are for vehicles in this decade. 
Points above the line are vehicles for which highway mpg > city mpg, points below are vehicles for which city mpg > highway mpg.

```{r}
## mpg_hawkes |>
##   filter(#...#) |>
##   ggplot(aes(#...#)) +
##   geom_point() + 
##   geom_abline(slope = 1, intercept = 0, lty = 2) # line of equality
```

# Problem: Plot Modifications

The following code makes the following adaptations to the visual you created in the previous problem. 

- Plot vehicles across all decades, rather than just the 1980s, coloring each point based on the decade that the vehicle was produced.     
HINT: Remember that the `floor()` function rounds down to the nearest whole number. Can you use that, division, and multiplication to make a decade variable? You may want to take the numeric decade (say you've called it `decade`) and use `decade = factor(decade, ordered = T)` to convert it to a factor so that `ggplot2` interprets it as a categorical variable. 
- Use the `geom_jitter()` geometry to randomly move the points slightly up/down and left/right to help with "overplotting" (multiple observations that have exactly the same x and y coordinates and are plotted on top of each other). 
- Use the `alpha` argument to make the points semi-transparent, which will further aid with the overplotting issue in the graph with so many observations. 
- Fit a "linear model" to the points to find the "line of best fit" between city and highway mileage and add it to the plot. This is the solid line. 
- Provide the "line of equality" as in the previous plot. This is the dashed line. 

```{r}
mpg_cars <- mpg_hawkes |>
  filter(v_class_2 == "Cars")

ggplot(mpg_cars, aes(x = city, y = highway, color = decade)) +
  geom_jitter(alpha = 0.5) + 
  geom_abline(intercept = 0, slope = 1, lty = 2) + # line of equality
  geom_smooth(aes(x = city, y = highway),
              inherit.aes = F, method = "lm", 
              color = "black") + # regression line
  coord_fixed()
```

Another way to do this is to explicitly fit the linear regression (this allows you to actually see what the equation is and/or examine how well it fits), and then plot the regression line directly. 

```{r}
mpg_lm <- lm(highway ~ city, data = mpg_cars)

ggplot(mpg_cars, aes(x = city, y = highway, color = decade)) +
  geom_jitter(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, lty = 2) + # line of equality
  geom_abline(intercept = mpg_lm$coefficients[1], # line of regression
              slope = mpg_lm$coefficients[2]) + 
  coord_fixed()
```

Notice in these plots that there are a lot of purple points that fall well below the line of best fit, which means that the highway mileage was less than expected based on their city mileage. **Discuss briefly (2 min or less) with a partner** what might cause modern vehicles to have much higher city mileage (as compared to highway mileage) than vehicles from previous decades. 

*Replace this text with your answer. Optional if discussing in a group and required if doing this on your own.*
