---
title: "5.2.a - Class Activity"
format: html
number-sections: true
execute:
  echo: true   # show code by default
  warning: false
  message: false
---

This assignment is intended to get you familiar with the tidyverse using real COVID-19 data. 

```{r}
library(tidyverse)
```


The following code chunk scrapes the `.csv` files we will be analyzing from the web renames and selects the subset of columns that we will be using for analysis. 

```{r}
url <- "https://srhdpeuwpubsa.blob.core.windows.net/whdh/COVID/"
url_case <- paste0(url, "WHO-COVID-19-global-table-data.csv")

covid_case <- read_csv(url_case) |>
  select(name = Name, region = `WHO Region`, 
         case_total = `Cases - cumulative total`,
         case_pop = `Cases - cumulative total per 100000 population`,
         death_total = `Deaths - cumulative total`,
         death_pop = `Deaths - cumulative total per 100000 population`)
```

The renamed columns that remain are: 

- `name`: Country name
- `region`: World Health Organization (WHO) region. 
- `case_total`: Cumulative number of reported COVID cases since Jan 2020
- `case_pop`: Cumulative number of reported COVID cases per 100,000 people since Jan 2020
- `death_total`: Cumulative number of reported COVID deaths since Jan 2020
- `death_pop`: Cumulative number of reported COVID deaths per 100,000 people since Jan 2020

# Countries in the Americas 

We are interested in comparing COVID statistics from the United States to other large countries in our region. Please create a new dataframe called `covid_case_am` which contains all countries in the "Americas" region. Use the pipe `|>` statement to help you complete the task. 

**How many countries are in this region?** Please answer this question in the associated canvas quiz. 

```{r}
## covid_case_am <- covid_case |> 
##   filter(#...#)
```

# Case Fatality Rate 

Within the "Americas" region, we want to know which countries had the highest percentage of deaths per reported case. This is calculated as $100 * \frac{\text{deaths\_total}}{\text{case\_total}}$. Create a new variable called `death_rate` that reflects this quantity. Then, after calculating this quantity, retain only countries whose death rate exceeds 3\% (or 3 deaths per 100 cases). Note that you do not need to save this dataframe to a variable, just show the output. 

**What are the three countries for which the death rate exceeds 3 deaths per 100 cases?** Answer this question in the associated canvas quiz. 

```{r}
## covid_case_am <- covid_case_am |>
##   mutate(#...#) 

## covid_case_am |>
##   filter(#...#)
```

# USA vs. Canada 

Using the variable you created in the previous problem, create a dataframe that includes only the "United States of America" and "Canada". Retain only the columns `name`, `case_pop`, `death_pop`, and `death_rate`.

**What is the difference in the deaths per 100,000 people, rounded to the nearest integer?** 

Please answer in your canvas quiz.  

```{r}
## covid_case_am |>
##   filter(#...#) |> 
##   select(#...#)
```

# Statistical Differences 

**Discuss with a partner:** What do you notice about the difference in the Canadian and American statistics? Comment briefly below. 

*Replace this text with your answer. Optional if you verbally discuss this question with a partner but required if you do this assignment on your own.*