---
title: "Pokemon `dplyr` practice"
code-annotations: hover
echo: true
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(readr) 
library(dplyr)
library(tidyr) 

poke <- read_csv("https://bit.ly/poke-data") # <1>
poke_types <- separate(poke, type, c("type1", "type2"), sep = ",") # <2>
```
1. Read in the CSV using the short bit.ly URL (to make it easier to follow along in class)
2. Create a data frame that splits type into two columns (before the comma, after the comma) -- this will be used for Your Turns.



## Filter

Using `poke_types`, get only columns matching any of the following criteria:

1. Pokedex number between 800 and 900
2. Grass type
3. Attack stats of more than 120


```{r}
#| echo: true
## Your code goes here ##
```

You should get only one pokemon: Rillaboom.

## Select 

Using the original `poke` data, get only columns matching any of the following criteria:


1. columns starting with `t`
2. the 9th column
3. the pokemon name

```{r}
#| echo: true
## Your code goes here ##
```

The output from this chunk should look roughly like this, but your row order may be different.

```
# A tibble: 1,526 × 4
   type          total attack name      
   <chr>         <dbl>  <dbl> <chr>     
 1 Grass           310     61 Sprigatito
 2 Fighting        455    105 Hitmonchan
 3 Fire            420     86 Raboot    
 4 Normal          420     70 Castform  
 5 Dark,Fighting   488     90 Scrafty   
 6 Dragon,Water    475     50 Tatsugiri 
 7 Psychic         600     95 Deoxys    
 8 Rock,Flying     500    100 Minior    
 9 Dragon,Ghost    410     80 Drakloak  
10 Water           295     40 Horsea    
# ℹ 1,516 more rows
# ℹ Use `print(n = ...)` to see more rows
```

## Mutate 

Create a column named `n_elevator` that calculates how many of these pokemon 
could fit in an elevator with a weight limit of 6000 lbs. 
You may assume that pokemon can be infinitely compressed, so that the volume of the pokemon is not a concern.

Round your answer down to the nearest integer with the `floor()` function - no one needs to get hurt!

Remember, you will need to convert kg to lbs, so you may want to create an additional column named `weight_lbs`.

Output only `gen`, `name`, and the columns you created.

```{r}
#| echo: true
## Your answer here ##
```

Your output should look something like this:

```
# A tibble: 1,526 × 5
     gen name       type          weight_lbs n_elevator
   <dbl> <chr>      <chr>              <dbl>      <dbl>
 1     9 Sprigatito Grass               9.04        663
 2     1 Hitmonchan Fighting          111.           54
 3     8 Raboot     Fire               19.8         302
 4     3 Castform   Normal              1.76       3401
 5     5 Scrafty    Dark,Fighting      66.1          90
 6     9 Tatsugiri  Dragon,Water       17.6         340
 7     3 Deoxys     Psychic           134.           44
 8     7 Minior     Rock,Flying        88.2          68
 9     8 Drakloak   Dragon,Ghost       24.3         247
10     1 Horsea     Water              17.6         340
# ℹ 1,516 more rows
# ℹ Use `print(n = ...)` to see more rows
```


## Group_by + Summarize! 

Work with `poke_types` and only worry about the primary type (`type1`)


### Bug Pokemon by Generation
Calculate the total number of bug pokemon in each generation.

```{r}
#| echo: true

## Your code here ##
```

Your output should be as follows:
```
# A tibble: 9 × 2
    gen   Bug
  <dbl> <int>
1     1    14
2     2    12
3     3    12
4     4    24
5     5    22
6     6     3
7     7     9
8     8     4
9     9     7
```


### Pokemon Weight By Type

Find the heaviest pokemon of each type, making sure to remove any duplicate rows. 

```{r}
## Your code here ##
```

Your results should be as follows:
```
# A tibble: 18 × 4
     gen name         type1    weight_kg
   <dbl> <chr>        <chr>        <dbl>
 1     3 Wailord      Water         398 
 2     6 Noivern      Flying         85 
 3     4 Giratina     Ghost         750 
 4     8 Eternatus    Poison        950 
 5     3 Groudon      Ground        950 
 6     7 Celesteela   Steel        1000.
 7     6 Xerneas      Fairy         215 
 8     7 Cosmoem      Psychic      1000.
 9     8 Glastrier    Ice           800 
10     4 Heatran      Fire          430 
11     1 Snorlax      Normal        460 
12     7 Guzzlord     Dark          888 
13     9 Iron Hands   Fighting      381.
14     9 Roaring Moon Dragon        380 
15     7 Stakataka    Rock          820 
16     7 Buzzwole     Bug           334.
17     9 Miraidon     Electric      240 
18     4 Torterra     Grass         310 
```

### Pokemon HP by Generation

Find the pokemon with the highest HP for each generation. Include all tied rows.

```{r}
## Your code here ##
```

```
# A tibble: 10 × 3
     gen name         hp
   <dbl> <chr>     <dbl>
 1     1 Chansey     250
 2     2 Blissey     255
 3     3 Wailord     170
 4     4 Giratina    150
 5     4 Drifblim    150
 6     5 Alomomola   165
 7     6 Zygarde     216
 8     7 Guzzlord    223
 9     8 Eternatus   255
10     9 Cetitan     170
```
