---
title: "Presidential Influence on the Country"
editor: source
format: 
  html:
    self-contained: true
---

# Motivation

We are constantly bombarded with stats and figures that shape our political opinions. This begs the question regarding best practices for interpreting such information. To hone that skill, this assignment asks you to find stats/tables/figures that support the following statements.

Please follow the instructions in the following subsections. 


# The United States is worse off as a result of the current president

## Value 

Please replace this text with the statement of value related to your stats/tables/figures. As a (hopefully) humorous example, suppose that I believe that there is a direct connection between the Joe Biden presidency and the happenings in the Pokemon Go community. My value might (humorously) be the following. 

*Pokemon Go fosters connections, community, and physical exercise. More Pokemon Go activity is a reflection of a society living more intentionally and more physically active.*

## Stat/Visual

Please provide a copy of your stat/table/visual in this section. Be sure also to provide a copy of the URL from which the statistics were obtained. Please ensure that your link is "clickable" before you provide it. See the example below. 

*Link to source: <https://prioridata.com/data/pokemon-go-stats/>*

Year	| Users (mm)
---- | ----
2016 |	232
2017 |	65
2018 |	131
2019 |	113
2020 |	86
2021 |	71
2022 |	62
2023 |	80
2024 |	106
2025 (Estimated) | 60 (Updated 4th Jan 2026)

: Table of annual active Pokemon Go Users, according to https://prioridata.com/data/pokemon-go-stats/


(I've created a line chart of the data shown on this webpage - it is fine for you just to show the table, or you can use AI to help you write the R code to create the chart. I just got a bit over-involved...)

```{r}
#| label: pokemon-go-data
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-alt: In 2016, there were 232 million pokemon go users; this dropped to 65 million in 2017, rose to 131 million in 2018, dropped to 113 million in 2019, 86 million in 2020, 71 million in 2021, 62 million in 2022, 55 million in 2023. A line chart showing these numbers is overlaid with blue (Democrat) and red (Republican) rectangles indicating who held the Oval office during each period.
#| fig-cap: Number of active Pokemon Go users from 2016-2023

library(tibble)

pokego <- tribble(~year, ~users_mil,
                  2016,232,
                  2017,65,
                  2018,131,
                  2019,113,
                  2020,86,
                  2021,71,
                  2022,62,
                  2023,80,
                  2024,106,
                  2025,60)
presidents <- tibble(year_start = c(2009, 2017, 2021, 2025)+1/12, year_end = c(2017, 2021, 2025, 2029)+1/12, president = c("Obama", "Trump", "Biden", "Trump"), party = c("blue", "red", "blue", "red"))

library(ggplot2)
ggplot(pokego, aes(x = year, y = users_mil)) + 
  geom_rect(data = presidents, aes(xmin = year_start, xmax = year_end, ymin = -Inf, ymax = Inf, fill = party), alpha = .1, inherit.aes = F) + 
  geom_line() + 
  geom_text(data = presidents, aes(x = year_end, y = 0, label = president), hjust = 1, inherit.aes = F) + 
  xlab("Year") + ylab("Users (million)") + ggtitle("Pokemon Go Users") + 
  coord_cartesian(xlim = c(2016, 2025), ylim = c(0, 250)) + scale_fill_identity() + 
  scale_x_continuous(breaks = seq(2016, 2026, by = 2)) + 
  theme_bw()
```


## Explanation of connection. 

Please replace this text with a description of how the above stat/visual supports your value statement. Please see the humorous example below. 

*The table shows a consistent decline of Pokemon Go users through most of Joe Biden's presidency. This suggests an inability of the Joe Biden Presidency to support and foster key community building activities such as Pokemon Go. As a result, our country is less connected and worse off as a result of Joe Biden's presidency.*


# The United States is better off as a result of the current president

## Value

Please replace this text with the statement of value related to your stats/tables/figures. As a (hopefully) humorous example, suppose that I believe that there is a direct connection between the Joe Biden presidency and the happenings in the Pokemon Go community. My value might (humorously) be the following.

**Pokemon Go user activity represents a breakdown of other, more meaningful social functions. Thus, increased Pokemon Go usage is a reflection of an inability for people to connect in more fulfilling ways.**

## Stat/Visual

Please provide a copy of your stat/table/visual in this section. Be sure also to provide a copy of the URL from which the statistics were obtained. Please ensure that your link is “clickable” before you provide it. See the example below.


*Link to source: <https://prioridata.com/data/pokemon-go-stats/>*

Year	| Users (mm)
---- | ----
2016 |	232
2017 |	65
2018 |	131
2019 |	113
2020 |	86
2021 |	71
2022 |	62
2023 |	80
2024 |	106
2025 (Estimated) | 60 (Updated 4th Jan 2026)

: Table of annual active Pokemon Go Users, according to https://prioridata.com/data/pokemon-go-stats/


(I've created a line chart of the data shown on this webpage - it is fine for you just to show the table, or you can use AI to help you write the R code to create the chart. I just got a bit over-involved...)

```{r}
#| label: pokemon-go-data
#| echo: false
```

## Explanation of Connection

Please replace this text with a description of how the above stat/visual supports your value statement. Please see the humorous example below.

**The table shows a consistent decline of Pokemon Go users through Joe Biden’s presidency. This suggests that other, more meaningful, social institutions must have been strengthened through the Joe Biden presidency, since the only reason someone would play Pokemon go is if they had no better alternative. This evidence of stronger social systems is evidence that our country is better off as a result of Joe Biden’s presidency.**